HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-24 docs/HLL_exp_bad/main.tex
tweaks

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Durand and Flajolet, 2003) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. We show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm, an adversary who has some flexibility in choosing what inputs will be recorded can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. 
\end{abstract}
}

@@

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own bucket. To map a value $x$ to one of $2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$ (so that the output length of $h(\cdot)$ is $n+\ell$). Estimators from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ empirically computed to correct a systematic multiplicative bias. As result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{add-ref-for-this-claim}.

An interesting property of HLL is that it supports merging in a lossless way. To combine two buckets at index $i$ in two different HLL instances, take the maximum of the two bucket entries and assign that to the matching bucket $i$ in the merged HLL. Such a simple set union operation allows easy parallelisation operations among multiple machines independently, provided they use the same hash function and the same number of buckets.

\subsection{Applications}
For completeness and motivation, we provide examples of concrete uses of HLL:
\begin{itemize}
    \item Cardinality estimation provides a significantly faster way for Facebook to compute statistics on their users. They use HLL to find out how many distinct people visited their website in the past week~\cite{fbhll}.
    \item HLL has been proposed as a solution to detecting Denial-of-Service attacks~\cite{portscanhll} (this paper actually uses a variation on HLL called sliding HLL~\cite{slidinghll} which estimates the cardinality over a moving time window). Specifically, HLL is used to count how many different ports are used as destination port over all packets received, as an attempt to identify when a port scan attack is underway.
        \item HLL has been implemented in network switches to approximate the number of distinct packets in traffic flows and overcome the switch resource limits~\cite{flexswitch}. The number of unique flows traversing a switch can then be used to provide a better congestion control protocol.
        \end{itemize}

@@

The work of~\cite{cardestprivacy} breaks the privacy of the HLL sketch and concludes that having the ability to test the insertion \kenny{what does this mean? see other comment later; the more I think about it, the more I think their paper is junk.} of a targeted item gives information about whether the item was originally present in the input data set. It is argued in~\cite{cardestprivacy} that HLL can be used by organizations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm.

@@

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as replacement (but at the cost of loosing mergability) or in addition to an unsalted sketch (inducing a memory overhead). 
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. This enables the adversary to control the output of hash function $h$ and flexibly choose which items will be inserted into the HLL, and in which buckets.

@@

\begin{itemize}
\item[S1:] the adversary has black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item; 
\item[S2:] the adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use.
\item[S3:] the adversary has direct access to the sketch in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL. Recall that these buckets hold the intermediate estimators that are averaged in a later step to compute the final cardinality estimate.
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln}. We do not put our focus on S1 as this previous work in~\cite{hllvuln} fully exploits this scenario. Their attack works as follow: ``after each insertion the cardinality is checked and the items which do not increment the HLL estimate are retained''. It is unclear what this means in practice, since after insertion into the HLL, an item cannot usually be removed by the adversary. \kenny{Mathilde: can you think about this sentence that I added? Do we need to rewrite S1? The scenario does not really make sense at all in practice. I think we might use this to beat up the other paper a bit more.}

@@

\noindent\paragraph{Scenario S3:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the intermediate estimators computed per bucket). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increment the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

\subsection{Impact}
We now evaluate the impact of the above attacks on the HLL cardinality estimate. We assume that $h$ is a ``good" hash function, so that the bits of $h(x)$ are balanced (i.e. have a probability of $1/2$ of equalling 0 and 1).

@@

In scenario S3, we compute the average work required of the adversary to find an input $x$ such that $h(x)$ meets the attack's requirements. Let us assume for the moment that $c_i \ge 1$ for all buckets $i$. Assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increment the estimator is the probability that $h(x)$'s rightmost $\ell$ bits have $c_i$ or less leading zeroes. This is equal to $1-\frac{1}{2^{c_i+1}}$ assuming $h$ behaves like a random function. Averaging over all $2^n$ buckets (and assuming the bucket choice is uniformly random), the probability that input $x$ does not increment the estimator is given by:
\begin{eqnarray*}
\sum_{i=1}^{2^n}\frac{1}{2^n} \cdot (1-\frac{1}{2^{c_i+1}}) & = & 1-\frac{1}{2}\cdot (\frac{1}{2^n}\cdot \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) \\
& = &1-(2H_c)^{-1} \\
\end{eqnarray*}
where $H_c$ is the harmonic mean of the counts $2^{c_1}, 2^{c_2}, \ldots, 2^{c_{2^n}}$.

%$P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i)=P(h(x)$ has LESS than $c_i$ leading 0s$)=1-P(h(x)$ has strictly MORE than $c_i$ leading 0s$)=1-\frac{1}{2^{c_i+1}}$\\
%Across all buckets, the adversary does not update any estimate with probability $P=
%\sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i$ AND $h(x)$ does NOT update bucket $i) = \sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i)P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i) = \sum_{i=1}^{2^n}\frac{1}{2^n} \times (1-\frac{1}{2^{c_i+1}}) = 1-\frac{1}{2}\times(\frac{1}{2^n}\times \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) = 1-\frac{1}{2}\times\frac{1}{H_c} = 1-(2H_c)^{-1}$, with $H_c$ the harmonic mean of the numbers $2^{c_1}, 2^{c_2}, ..., 2^{c_{2^n}}$.
So the adversary needs to craft on average $(1-(2H_c)^{-1})^{-1}$ distinct candidates for $x$ in order to insert a single one, and, given $t$ bits of flexibility in its choice of inputs, can expect to ``hide" up to $(1-(2H_c)^{-1}) \cdot 2^t$ distinct items without increasing the HLL cardinality estimate at all.

@@

Note that the above analysis made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are zero, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$.

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Durand and Flajolet, 2003) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm, an adversary who has some flexibility in choosing what inputs will be recorded can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. 
\end{abstract}
}

@@

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own bucket. The collection of buckets is referred to as the HLL \emph{sketch}. To map a value $x$ to one of $2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$ (so that the output length of $h(\cdot)$ is $n+\ell$). Estimators from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ empirically computed to correct a systematic multiplicative bias. As result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{add-ref-for-this-claim}.

An interesting property of HLL is that it supports merging in a lossless way. To combine two buckets at index $i$ in two different HLL instances, take the maximum of the two bucket entries and assign that to the matching bucket $i$ in the merged HLL sketch. Such a simple set union operation allows easy parallelisation of operations among multiple machines independently, provided they use the same hash function and the same number of buckets.

\subsection{Applications}
For completeness and motivation, we provide examples of concrete uses of HLL:
\begin{itemize}
    \item Cardinality estimation provides a significantly faster way for Facebook to compute statistics on their users. They use HLL to find out how many distinct people visited their website in the past week~\cite{fbhll}.
    \item HLL has been proposed as a solution to detecting Denial-of-Service attacks~\cite{portscanhll} (this paper actually uses a variation on HLL called \emph{sliding HLL}~\cite{slidinghll} which estimates the cardinality over a moving time window). Specifically, HLL is used to count how many different ports are used as destination port over all packets received, as an attempt to identify when a port scan attack is underway.
        \item HLL has been implemented in network switches to approximate the number of distinct packets in traffic flows and overcome the switch resource limits~\cite{flexswitch}. The number of unique flows traversing a switch can then be used to provide a better congestion control protocol.
        \end{itemize}

@@

The work of~\cite{cardestprivacy} breaks the privacy of HLL and concludes that having the ability to test the insertion \kenny{what does this mean? see other comment later} of a targeted item gives information about whether the item was originally present in the input data set. It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm.

@@

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as replacement (but at the cost of losing mergability) or in addition to an unsalted sketch (inducing a memory overhead). 
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. This enables the adversary to control the output of hash function $h$ and flexibly choose which items will be inserted into the HLL sketch, and in which buckets.

@@

\begin{itemize}
\item[S1:] the adversary has black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item; 
\item[S2:] the adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use.
\item[S3:] the adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch. Recall that these buckets hold the intermediate estimators that are averaged in a later step to compute the final cardinality estimate.
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln}. We do not put our focus on S1 as this previous work in~\cite{hllvuln} fully exploits this scenario. Their attack works as follow: ``after each insertion the cardinality is checked and the items which do not increment the HLL estimate are retained''. It is unclear what this means in practice, since after insertion into the HLL, an item cannot usually be removed by the adversary. \kenny{Mathilde: can you think about this sentence that I added? Do we need to rewrite S1? The scenario does not really make sense at all in practice. I think we might use this to beat up the other paper a bit more. Maybe it also applies to the ``privacy'' paper.}

@@

\noindent\paragraph{Scenario S3:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the count values per bucket in the HLL sketch). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increment the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

\subsection{Impact}
We now evaluate the impact of the above attacks on the HLL cardinality estimate. We assume that $h$ is a ``good" hash function, so that each of the bits of $h(x)$ can be regarded as being independent and uniformly random.

@@

In scenario S3, we compute the average work required of the adversary to find an input $x$ such that $h(x)$ meets the attack's requirements. Let us assume for the moment that $c_i \ge 1$ for all buckets $i$. Assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increment the estimator is the probability that $h(x)$'s rightmost $\ell$ bits have $c_i$ or less leading zeroes; this probability is equal to $1-\frac{1}{2^{c_i+1}}$ assuming $h$ behaves like a random function. Averaging over all $2^n$ buckets (and using the fact that the bucket choice is uniformly random since the bits of $h$ are independent and uniformly random), the probability that input $x$ does not increment the estimator is given by:
\begin{eqnarray*}
\sum_{i=1}^{2^n}\frac{1}{2^n} \cdot (1-\frac{1}{2^{c_i+1}}) & = & 1-\frac{1}{2}\cdot (\frac{1}{2^n}\cdot \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) \\
& = &1-(2H_c)^{-1} \\
\end{eqnarray*}
where $H_c$ is the harmonic mean of the counts $2^{c_1}, 2^{c_2}, \ldots, 2^{c_{2^n}}$.

%$P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i)=P(h(x)$ has LESS than $c_i$ leading 0s$)=1-P(h(x)$ has strictly MORE than $c_i$ leading 0s$)=1-\frac{1}{2^{c_i+1}}$\\
%Across all buckets, the adversary does not update any estimate with probability $P=
%\sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i$ AND $h(x)$ does NOT update bucket $i) = \sum_{i=1}^{2^n}P(h(x)$ mapped to bucket $i)P(h(x)$ does NOT update bucket $i| h(x)$ mapped to bucket $i) = \sum_{i=1}^{2^n}\frac{1}{2^n} \times (1-\frac{1}{2^{c_i+1}}) = 1-\frac{1}{2}\times(\frac{1}{2^n}\times \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) = 1-\frac{1}{2}\times\frac{1}{H_c} = 1-(2H_c)^{-1}$, with $H_c$ the harmonic mean of the numbers $2^{c_1}, 2^{c_2}, ..., 2^{c_{2^n}}$.
So the adversary needs to craft on average $(1-(2H_c)^{-1})^{-1}$ distinct candidates for $x$ in order to insert a single candidate, and, given $t$ bits of flexibility in its choice of inputs, can expect to ``hide" up to $(1-(2H_c)^{-1}) \cdot 2^t$ distinct items without increasing the HLL cardinality estimate at all.

@@

Note that the above analysis made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are 0, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$.
